Joint training of classification model and similarity model for low-resource text classification in chatbot

### 2020-08-06 chatbot_classification/aaai_2021/paper.tex
update

\section{Methodology}
In this section, we will describe a system of similarity model fused with classification model, named as SFC, in detail. We come up with 2 different approaches for implementation of SFC system, in which hierarchical SFC is a further improvement of 2-stage SFC for better performance. 
\subsection{2-stage SFC with multi-task learning}
The detailed network structure of 2-stage SFC is shown in Fig. \ref{fig:ver.1.3}. It is composed with 2 separate stages in charge of different jobs.
In stage 1, we can use any auxiliary model or tool to provide top-K most related classes for sampling sentence pairs of both positive samples and negative samples for stage 2. Suppose we have a training data set $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$ of $N$ data points, in which $x_{i}$ is the user input query and $y_{i}$ is a single class label from a class label set of $C$ classes in total. If we want to generate sentence pair data set $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$, where $(x, x')_{j}$ is sentence pair grouped by 2 user input queries and $l_{j}\in\{0, 1\}$ is similarity class label, see Equ. \ref{eq:similarity label}, 
\begin{align}
l_{j}=\begin{cases} \label{eq:similarity label}
0 & y_{s_1}\neq y_{s_2}\\
1 & y_{s_1}=y_{s_2}
\end{cases}
\end{align}
directly from original data set $\mathcal{D}$ without any sampling strategy, then $\mathcal{D'}$ will have $M=N^2$ data points of sentence pairs, which might be too time-consuming for the training process of stage 2 since we often have hundreds of class labels under task-specific chatbot scenario. Although we may only have a few data points for a single class label, $N$ is still quite a big number.

Therefore, we adopt the idea of adversarial negative sampling strategy raised by \cite{bamler2020extreme}. The key idea of this sampling strategy is to train with negative samples that are hard to distinguish from positive samples. Now suppose we have a user input query $x_{i}$ as $s_{1}$ and top-K most related class labels set $\mathcal{C'}=\{c'_{1}, \dots, c'_{K}\}$ found by auxiliary model or tool in stage 1, where $K$ is a hyperparameter set by us that controls candidate class number, we apply the adversarial negative sampling strategy into our sentence pair sampling process by the following 2 steps:

@@

\textbf{2. negative sentence pair sampling}: As for negative sentence pairs, we will also randomly sample $P$ sentences for each class in the negative candidate class label set $\mathcal{C'}\backslash \{y_{i}\}$. The class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are the set of most confusing classes comparing to the ground truth ${y_{i}}$, so we assume that the sentence pairs with negative label grouped by user input query $x_{i}$ and sentences with class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are also strong adversarial negative samples that can help enhance the training speed and performance. According to the same method as positive sentence pair sampling, we can obtain the set of sentence pairs with negative label as $\mathcal{N'}_{i}=\{[(x_{i}, x'_{0}), 0], [(x_{i}, x'_{1}), 0], ..., [(x_{i}, x'_{P\cdot (K-1)}), 0]\}$

@@

\subsubsection{stage 1: classification model for providing top-K classes}
As for stage 1, our SFC system can adapt to any auxiliary model or even searching tool based on Term Frequency-Inverse
Document Frequency (TF-IDF) retrieval. In this paper, however, we choose to fine-tune RoBERTa on our task-specific data set $\mathcal{D}$ by adding a softmax classifier on top of it, the same as the structure shown in Fig. \ref{Fig.1.sub.1}, for better overall performance. According to our experimental result, RoBERTa can guarantee that approximately 95\% ground truth label be included in top-5 candidate classes with highest probability after being fine-tuned on task-specific data set with top-1 weighted F score at around 0.75. So we use RoBERTa as encoder of classification model in our setting at stage 1 to ensure low limitation for the performance of stage 2.

@@

\subsubsection{stage 2: sentence-pair model for multi-task learning}
As for stage 2, we also choose RoBERTa as the encoder for sentence pairs, as shown in Fig. \ref{fig:ver.1.3}. Before sending sentence pairs to Roberta, we pack sentence pairs into a single sequence and separate them with a special token [SEP]. Roberta will also use a learned embedding to every token indicating whether it belongs to the first sentence or the second one. Then, we utilize the cross-attention mechanism inside RoBERTa, and then extract semantic relationship between 2 sentences from the output softmax layer, which is placed on top of the last hidden representation of the [CLS] token.

@@

Therefore, to further improve the overall performance of SFC, we proposed a hierarchical SFC joint model structure, shown in Fig. \ref{fig:framework}. In hierarchical SFC, classification model is being placed in lower level to dynamically provide top-K candidate classes for the sentence-pair similarity model placed in higher level through a top-K pooling layer. In this way, classification model and similarity model can be merged into one single joint-model for multi-task training, thus avoiding the limitation brought by 2 separate stage structure.

@@

@@

>>>>

\section{Methodology}
In this section, we will describe a system of similarity model fused with classification model, named as SFC, in detail. We come up with 2 different approaches for implementation of SFC system, in which hierarchical SFC is a further improvement of 2-stage SFC for better performance. 
\subsection{2-stage SFC with multi-task learning}
The detailed network structure of 2-stage SFC is shown in Fig. \ref{fig:ver.1.3}. It is composed with 2 separate stages in charge of different jobs.
In stage 1, we can use any auxiliary model or tool to provide top-K most related classes for sampling sentence pairs of both positive samples and negative samples for stage 2. Suppose we have a training data set $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$ of $N$ data points, in which $x_{i}$ is the user input query and $y_{i}$ is a single class label from a class label set of $C$ classes in total. If we want to generate sentence pair data set $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$, where $(x, x')_{j}$ is sentence pair grouped by 2 user input queries and $l_{j}\in\{0, 1\}$ is similarity class label, see Equ. \ref{eq:similarity label}, 
\begin{align}
l_{j}=\begin{cases} \label{eq:similarity label}
0 & y_{x}\neq y_{x'}\\
1 & y_{x}=y_{x'}
\end{cases}
\end{align}
directly from original data set $\mathcal{D}$ without any sampling strategy, then $\mathcal{D'}$ will have $M=N^2$ data points of sentence pairs, which might be too time-consuming for the training process of stage 2 since we often have hundreds of class labels under task-specific chatbot scenario. Although we may only have a few data points for a single class label, $N$ is still quite a big number.

Therefore, we adopt the idea of adversarial negative sampling strategy raised by \cite{bamler2020extreme}. The key idea of this sampling strategy is to train with negative samples that are hard to be distinguished from positive samples. Now suppose we have a user input query $x_{i}$ and top-K most related class labels set $\mathcal{C'}=\{c'_{1}, \dots, c'_{K}\}$ found by auxiliary model or tool in stage 1, where $K$ is a hyperparameter set by us that controls candidate class number, we apply the adversarial negative sampling strategy into our sentence pair sampling process by the following 2 steps:

@@

\textbf{2. negative sentence pair sampling}: As for negative sentence pairs, we will also randomly sample $P$ sentences for each class in the negative candidate class label set $\mathcal{C'}\backslash \{y_{i}\}$. The class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are the set of most confusing classes comparing to the ground truth ${y_{i}}$, so we assume that the sentence pairs with negative label grouped by user input query $x_{i}$ and sentences with class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are also strong adversarial negative samples for sentence-pair similarity model that can help enhance the training speed and performance. According to the same method as positive sentence pair sampling, we can obtain the set of sentence pairs with negative label as $\mathcal{N'}_{i}=\{[(x_{i}, x'_{0}), 0], [(x_{i}, x'_{1}), 0], ..., [(x_{i}, x'_{P\cdot (K-1)}), 0]\}$

@@

\subsubsection{stage 1: classification model for providing top-K classes}
As for stage 1, our SFC system can adapt to any auxiliary model or even searching tool based on Term Frequency-Inverse
Document Frequency (TF-IDF) retrieval. In this paper, however, we choose to fine-tune RoBERTa on our task-specific data set $\mathcal{D}$ by adding a softmax classifier on top of it, the same as the structure shown in Fig. \ref{Fig.1.sub.1}, for better overall performance. According to our experimental result, RoBERTa can guarantee that approximately 95\% ground truth label be included in top-5 candidate classes with highest probability after being fine-tuned on task-specific data set with experimental result measured by F-score at around 0.75. So we use RoBERTa as encoder of classification model in our setting at stage 1 to ensure low limitation for the performance of stage 2.

@@

\subsubsection{stage 2: sentence-pair model for multi-task learning}
As for stage 2, we also choose RoBERTa as the encoder for sentence pairs, as shown in Fig. \ref{fig:ver.1.3}. Before sending sentence pairs to Roberta, we pack sentence pairs into a single sequence and separate them with a special token [SEP]. Roberta will also add a learned embedding to every token indicating whether it belongs to the first sentence or the second one. Then, we utilize the cross-attention mechanism inside RoBERTa, and then extract semantic relationship between 2 sentences from the output softmax layer, which is placed on top of the last hidden representation of the [CLS] token.

@@

Therefore, to further improve the overall performance of SFC, we proposed a hierarchical joint model structure, shown in Fig. \ref{fig:framework}. In hierarchical SFC, classification model is being placed in lower layer level to dynamically provide top-K candidate classes for the sentence-pair similarity model placed in higher layer level through a top-K pooling layer. In this way, classification model and similarity model can be merged into one single joint-model for multi-task training, thus avoiding the limitation brought by 2 separate stage structure.

@@

\textbf{Amazon-670K} is a customer product review dataset for text classification task from the extreme classification repository with full data containing 670,091 class labels, 285,176 training samples and 150875 test samples\cite{bhatia2016extreme}. As the data set contains multiple class labels per training sample, we keep only the first class label for each data point. Besides, we also randomly sampled 250 class labels from all the classes that contain only 5-15 data points. In this way, we simulate the low-resource chatbot corpus setting with the subset of 1,942 training samples and 716 test samples for 250 class labels.

$\bullet$ \textbf{subset of recently-released intent detection datasets}.

From our point of view, task-specific chatbot usually starts with small amount of data at the initial stage. As more and more customers start to communicate with the chatbot, more data can be extracted from conversation log. Base on these assumptions, we do experiments on the following 3 intent detection dataset under sampling method similar to the approach in \cite{casanueva2020efficient} but in more sophisticated few-shot settings, while keeping the same test sets for each experiment to see the change of the improvement scale with decreasing data amount.

\textbf{HWU64} is a intent detection dataset designed for home robot scenario\cite{liu2019benchmarking}. It aims at the specific task of capturing the intent for different user requests to home robot and finding the corresponding answer. The raw dataset contains 25,716 data points for 64 class labels through crowdsourcing, which raises much more data than common real-world circumstances. Therefore, we experiment with the setups where only 5, 10, 15 and 20 training data points are available for each class label by random sampling.

\textbf{CLINC150} is a dataset designed for task-oriented systems with 23,700 queries that are short and unstructured for 150 intents, in the same style made by real users through crowdsourcing\cite{larson2019evaluation}. Here, we sample 5, 10 and 15 training data points for each intent class label by random sampling for our experimental setup.

\textbf{BANKING77} is a intent detection dataset for bank customer services. The raw dataset contains 13,084 data points for 77 class labels\cite{casanueva2020efficient}. Here, we sample 5, 10, 15, 20 training data points for each class label by random sampling for our experimental setup.

\begin{table*}
\begin{centering}
\scalebox{0.8}{
\begin{tabular}{|c|ccc|cccc|cccc|c|c|}
\hline 
\multicolumn{1}{|c|}{\multirow{2}*{\textbf{Models}} }& \multicolumn{3}{c|}{\textbf{CLINC150}}& \multicolumn{4}{c|}{\textbf{BANKING77}}& \multicolumn{4}{c|}{\textbf{HWU64}}& \ \ \ \textbf{ITG} \ \ \ & \textbf{Amazon-670k}\tabularnewline
\cline{2-14}
\multicolumn{1}{|c|}{} & $\textbf{5}$ & $\textbf{10}$ & $\textbf{15}$ & $\textbf{5}$ & $\textbf{10}$ & $\textbf{15}$ & $\textbf{20}$ & $\textbf{5}$ & $\textbf{10}$ & $\textbf{15}$ & $\textbf{20}$ & \textbf{Full} & \textbf{Full}\tabularnewline
\hline
BERT-large & \multirow{2}*{$0.8080$} & \multirow{2}*{$0.8904$} & \multirow{2}*{$0.9265$} & \multirow{2}*{$0.578$} & \multirow{2}*{$0.8004$} & \multirow{2}*{$0.8518$} & \multirow{2}*{$0.8827$} & \multirow{2}*{$0.4711$} & \multirow{2}*{$0.5963$} & \multirow{2}*{$0.6342$} & \multirow{2}*{$0.7010$} & \multirow{2}*{$0.7485$} & \multirow{2}*{$0.6658$}\tabularnewline
(classification) & & & & & & & & & & & & &\tabularnewline
\hline
ALBERT-xxlarge & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-}\tabularnewline
(classification) & & & & & & & & & & & & &\tabularnewline
\hline
RoBERTa-base & \multirow{2}*{$0.8732$} & \multirow{2}*{$0.9254$} & \multirow{2}*{$0.9363$} & \multirow{2}*{$0.7305$} & \multirow{2}*{$0.8654$} & \multirow{2}*{$0.8808$} & \multirow{2}*{$0.9080$} & \multirow{2}*{$0.5831$} & \multirow{2}*{$0.6790$} & \multirow{2}*{$0.7064$} & \multirow{2}*{$0.7100$} & \multirow{2}*{$0.7734$} & \multirow{2}*{$0.6708$}\tabularnewline
(classification) & & & & & & & & & & & & &\tabularnewline
\hline
RoBERTa-large & \multirow{2}*{$0.8974$} & \multirow{2}*{$0.9372$} & \multirow{2}*{$0.9508$} & \multirow{2}*{$0.7690$} & \multirow{2}*{$0.8728$} & \multirow{2}*{$0.8966$} & \multirow{2}*{$0.9099$} & \multirow{2}*{$0.6044$} & \multirow{2}*{$0.7002$} & \multirow{2}*{$0.7129$} & \multirow{2}*{$0.7436$} & \multirow{2}*{$0.7990$} & \multirow{2}*{$0.7156$}\tabularnewline
(classification) & & & & & & & & & & & & &\tabularnewline
\hline
RoBERTa-large & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-}\tabularnewline
(similarity) & & & & & & & & & & & & &\tabularnewline
\hline
2-stage SFC & \multirow{2}*{$0.8979$} & \multirow{2}*{$0.9457$} & \multirow{2}*{$0.9517$} & \multirow{2}*{$0.7975$} & \multirow{2}*{$0.8818$} & \multirow{2}*{$0.8962$} & \multirow{2}*{$0.9109$} & \multirow{2}*{$0.6477$} & \multirow{2}*{$0.7055$} & \multirow{2}*{$0.7200$} & \multirow{2}*{$0.7232$} & \multirow{2}*{$0.7972$} & \multirow{2}*{$0.7189$}\tabularnewline
(task1 only) & & & & & & & & & & & & &\tabularnewline
\hline
2-stage SFC & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-}\tabularnewline
(task2 only) & & & & & & & & & & & & &\tabularnewline
\hline
2-stage SFC & \multirow{2}*{$0.9167$} & \multirow{2}*{$0.9456$} & \multirow{2}*{$0.9571$} & \multirow{2}*{$0.7954$} & \multirow{2}*{$0.8854$} & \multirow{2}*{$0.8931$} & \multirow{2}*{$0.9192$} & \multirow{2}*{$0.6525$} & \multirow{2}*{$0.7001$} & \multirow{2}*{$0.7168$} & \multirow{2}*{$0.7476$} & \multirow{2}*{$\textbf{0.8124}$} & \multirow{2}*{$0.7364$}\tabularnewline
(task1 + task2) & & & & & & & & & & & & &\tabularnewline
\hline
\multirow{2}*{Hierarchical SFC} & \multirow{2}*{$\textbf{0.9231}$} & \multirow{2}*{$\textbf{0.9560}$} & \multirow{2}*{$\textbf{0.9644}$} & \multirow{2}*{$\textbf{0.8270}$} & \multirow{2}*{$\textbf{0.9069}$} & \multirow{2}*{$\textbf{0.9103}$} & \multirow{2}*{$\textbf{0.9209}$} & \multirow{2}*{$\textbf{0.6697}$} & \multirow{2}*{$\textbf{0.7211}$} & \multirow{2}*{$\textbf{0.7254}$} & \multirow{2}*{$\textbf{0.7497}$} & \multirow{2}*{$0.8114$} & \multirow{2}*{$\textbf{0.7445}$}\tabularnewline
 & & & & & & & & & & & & &\tabularnewline
\hline
\end{tabular}
}
\par\end{centering}
\caption{F-scores on five task-specific dataset for text classification in chatbot under low resource. For ITG, we keep the full dataset. For Amazon-670k, we randomly sampled 250 classes with training sample numbers within 5-15 samples per class. For CLINC150, BANKING77, HWU64, we set up various few-shot settings (5/10/15/20 samples per class) while keeping the test set to be fixed. The highest scores for each data setting are marked in bold.}

\label{tbe:table2}
\end{table*}

\subsection{Baselines}
We choose 3 groups of representative baselines:

\textbf{Group 1: single-task classification model.} This group of baseline models are based on the network structure for classification model shown in \ref{Fig.1.sub.1}. As for the pretrained model, we choose RoBERTa, BERT and ALBERT, all of which are among the most powerful pretrained models that achieve outstanding results in many NLP tasks, in our experimental setting. Since we apply RoBERTa in our SFC system, we naturally adopt single-task classification model based on RoBERTa-large and RoBERTa-base as our baseline models. Besides, we also adopt the baseline pre-trained model BERT-large\cite{devlin2018bert} and recently released ALBERT-xxlarge\cite{lan2019albert} for comparison. Since we use different few-shot settings from \cite{casanueva2020efficient} for those intent detection datasets, we also use BERT-large tuned on our few-shot settings as an essential reference for experimental results.

\textbf{Group 2: single-task similarity model.} This group of baseline models are based on network structure for similarity model shown in \ref{Fig.1.sub.2}. We choose RoBERTa-large for the pretrained model part as an powerful baseline. But during the evaluation step, we use elastic search to do a rough filter for candidate sentence pairs to guarantee a reasonable inference time for each user query.

\textbf{Group 3: SFC system with ablation study.} This group of baseline is based on the network structure of 2-stage SFC shown in \ref{fig:ver.1.3}. We conduct ablation study by introducing single task SFC system with only Task 1 or Task 2 to investigate the influence of each component within the total loss.

\subsection{Implementation Details}
For our SFC system, we use RoBERTa-large as the pretrained model for both classification task and sentence-pair similarity task. We further fine-tune our SFC system on task-specific datasets with max sequence length of 128, learning rate of 1.5e-5 for RoBERTa-large pretrained model and learning rate of 5e-4 for task-specific layer with polynomial decay and Adam optimizer. Also, we set the max number of the epoch to 100 and save the best model on validation set based on test result in F-score. We also report the F-score as the main evaluation measure for all experimental results.

For the experiments in Table \ref{tbe:table2}, we empirically set the hyperparameter to be $K=5$ and $P=10$ for all SFC variants. In Table \ref{tbe:table3}, we conduct extensive experiments to see the performance of SFC system under various hyperparameter settings.

\subsection{Result Analysis}

Table \ref{tbe:table2} presents the results for each experimental data and model set-up in detail. The performance of each model is measured in F-score on the fixed test set for each data set. The overall performance of SFC system is better than all the competitive single-model baseline models on five experimental datasets. This observation gives us a general idea that classification model and similarity model can actually provide each other with complementary information, and combining them together into a joint system can effectively enhance the overall capability based on richer extracted semantic information. However the relative performance among all the variants of SFC also seems to depend on some specific charateristic of each data set. Therefore, we did the following more detailed analysis from several different aspect.

@@

\begin{tabular}{|c|c|c|c|c|c|c|}
\hline 
\multicolumn{1}{|c|}{\multirow{2}*{Dataset}}&\multicolumn{1}{c|}{\multirow{2}*{Model}}& $K=3$ & $K=5$ & $K=10$ & $K=15$ & $K=20$\tabularnewline
\multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{} & $P=20$ & $P=10$ & $P=5$ & $P=4$ & $P=3$\tabularnewline
\hline
\multicolumn{1}{|c|}{\multirow{2}*{ITG}}& 2-stage SFC & 0.05 & 0.05 & 0.05 & 0.05 & 0.05\tabularnewline
\cline{2-7}
\multicolumn{1}{|c|}{} & Hierarchical SFC & 0.05 & 0.05 & 0.05 & 0.05 & 0.05\tabularnewline
\hline
\multicolumn{1}{|c|}{\multirow{2}*{Amazon-670k}}& 2-stage SFC & 0.05 & 0.05 & 0.05 & 0.05 & 0.05\tabularnewline
\cline{2-7}
\multicolumn{1}{|c|}{} & Hierarchical SFC & 0.05 & 0.05 & 0.05 & 0.05 & 0.05\tabularnewline
\hline
\end{tabular}
\par\end{centering}
\caption{We show the performance of SFC from different settings of hyperparameters for $K$, the candidate class number, and $P$, the sample number for each class.}

\label{tbe:table3}
\end{table*}
